### 2018-03-14 paper/3-abstractions.tex
Work on S3.1-S3.3

\begin{figure}
\begin{minted}{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- State monad
data State s
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of library functions}\label{fig-lib-funs}
\vspace{-4mm}
\end{figure}

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up to date. The store is a key-value map, together with \emph{persistent build
information} that is maintained by the build system itself.

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.
Fig.~\ref{fig-lib-funs} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells). This, in particular, suggests that it is possible to
\emph{depend on the absence of a value}, which is a useful feature for build
systems. In addition to usual \hs{getValue} and \hs{putValue} operations, some
stores also support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}. For example, GVFS
(Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
contents only on demand. The store also contains a single persistent value, of
type \hs{i}, representing the persistent state maintained by the build system
itself~--~its~``memory''. We can \hs{initialise} a store by supplying an initial
value \hs{i} and a key-value map \store.

@@

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
a task description (in the \Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{fetch} to find the value of any keys on
which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures
are given in Fig.~\ref{fig-lib-funs}. We will return to the need for this
generality in~\S\ref{sec-deps}.

For now, we content ourselves with observing that a task description,
of type \hs{Task c k v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v
busy task key store = execState (go key) store
  where
    go k = case task go k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\emph{SLPJ: explain}

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- State monad
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up-to-date. The store is a key-value map, together with \emph{persistent build
information} that is maintained by the build system itself.

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells). In addition to usual \hs{getValue} and \hs{putValue} operations,
some stores support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}\footnote{For
example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
file contents only on demand.}. The store also holds a value of type \hs{i},
representing the persistent build information maintained by the build
system~--~its ``memory''. Given an \hs{i} and a key-value map \store, we can
\hs{initialise} a store.

% This, in particular, suggests that it is possible to \emph{depend on the
% absence of a value}, which is a useful feature for build systems.

@@

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
return to the need for this generality in~\S\ref{sec-deps}.

For now, we content ourselves with observing that a task description,
of type \hs{Task}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date. What exactly does ``up-to-date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v
busy task key store = execState (compute key) store
  where
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system uses a function
\hs{compute}~\hs{::}~\hs{k}~\hs{->}~\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)}~\hs{v},
which given a key brings the associated value up-to-date and returns it. To do
so, \hs{compute} calls \hs{task} passing itself as an argument, so that \hs{task}
can use \hs{compute} to recursively find the values of its dependencies. If
\hs{task} returns \hs{Nothing} (this key is an input), \hs{compute} simply reads
the result from the store~\hs{s}, which is available in the \hs{State} monad (in
which \hs{compute} runs) via the \hs{get} method~--~see the \hs{State} monad
interface in Fig.~\ref{fig-types}. Otherwise it runs the action \hs{act}
returned by the \hs{task} to produce a resulting value~\hs{v}, stores it
(modifying the store \hs{s}), and returns it.

Given an acyclic task description, the \hs{busy} build system terminates with a
correct result, but it is not a minimal build system (c.f. the
Definition~\ref{def-minimal}); \hs{busy} keeps no track of keys it has already
brought up-to-date, and will therefore busily recompute the same keys again and
again if they have multiple dependents. We will develop much more efficient
build systems in~\S\ref{sec-implementations}.

### 2018-03-14 paper/4-build.tex
Rename section file
Work on S3.1-S3.3

The topological approach creates a linear ordering, which when followed, ensures each build is correct. Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort. However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{getdependencies}.

@@

We assume that \hs{Hash v} is a small constant size, constructed from hashing the underlying \hs{v} rather than storing it directly. Checking a trace requires ensuring all the dependencies are up to date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same. After computing a fresh value we add its \hs{Trace} to the list. Therefore the information stored by a build system that verifies traces is \hs{[Trace k v]}.

>>>>

The topological approach creates a linear ordering, which when followed, ensures each build is correct. Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort. However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-deps}.

@@

We assume that \hs{Hash v} is a small constant size, constructed from hashing the underlying \hs{v} rather than storing it directly. Checking a trace requires ensuring all the dependencies are up-to-date (using whatever ordering strategy as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are same as the current value and that the result is the same. After computing a fresh value we add its \hs{Trace} to the list. Therefore the information stored by a build system that verifies traces is \hs{[Trace k v]}.
